Plant Metagenomic Barcoding of Pollen Loads Offers Insights on the Foraging

### 2022-11-20 script/manuscript_MEE.Rmd
updated some text and style

---
title: Plant Metagenomic Barcoding of Pollen Loads Offers Insights on the Foraging
  Patterns of Queen Bumble Bees in the Southern Rocky Mountains, U.S.A.
author: |
  | Reed Clark Benkendorf^[Correspondence: reedbenkendorf2021@u.northwestern.edu] $^1$$^,$$^2$, Jane E. Ogilvie $^3$, Emily J. Woodworth $^1$$^,$$^2$, Sophie Taddeo $^1$$^,$$^2$, Paul J. CaraDonna $^1$$^,$$^2$$^,$$^3$, Jeremie B. Fant $^1$$^,$$^2$ | $^1$Chicago Botanic Garden, 1000 Lake Cook Road, Glencoe, Illinois 60022 USA,  $^2$Plant Biology and Conservation, Northwestern University, Evanston, IL 60208, USA, $^3$ Rocky Mountain Biological Laboratory, P.O. Box 519, Crested Butte, CO 81224, USA
abstract: |
  an abstract will be written to fill this space.
output:
  pdf_document: default
  toc: no
csl: ../citations/methods-in-ecology-and-evolution.csl
bibliography: ../citations/citations.bib
link-citations: yes
fig_caption: yes
header-includes:
- \usepackage{endfloat}
- \usepackage{setspace}\doublespacing
- \usepackage{lineno}
- \linenumbers
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```

@@

# INTRODUCTION
The inability to reliably identify plants to the level of species often leaves our understanding of ecosystem function and interactions nebulous. Current methods to remediate this situation include: ignoring these ecologically relevant levels of detail, revisiting plots for diagnostic material, assistance from taxonomic specialists, or the use of barcoding or other molecular techniques. These approaches are untenable in light of the benefits offered by species in several complex genera which serve as bioindicators, and preferred partners in ecological interactions, as well as an increasing lack of taxonomic experts. Many genera have species which are well defined based upon ecological rather than morphological properties, the identification of these taxa in degraded areas or without their mutualistic partners is fraught with difficulty, hindering an understanding of the breadth of habitat which some species occupy, and the interactions they have with other species. 
  The identification of many plant species to terminal taxon is an essential component of nearly all land management programs, where many species in the same genus  (e.g. Sagebrush - *Artemisia* L., Willows - *Salix* L., and Sedges - *Carex* L.) serve as bioindicators, as well as in academic research (@Gage2013HistoricalRO, AIM). This endeavour is often mired by lack of diagnostic characters (e.g. flowers, fruits, or roots),  and increasingly the description of cryptic species (@janzen2017nuclear, @oliver2009cryptic). Solutions to this problem are wanting, certain programmes have relied increasingly upon revisiting field sites to identify material using morphological or chemical approaches, whereas academic research has often used high copy number plastid genes as barcodes (Rosentreter et al. 2021, MORE MORE). However, both approaches have significant downsides, the former resource intensive at the landscape scale - and often does not work, while the latter seldom works due to a lack of variability in the currently available barcodes (@liu2014identification).
  Recently barcoding, and metabarcoding, have shown much promise in all Kingdoms of life. For example …. . With plants the identification of members of certain clades has been quite successful, whereas with others results have been elusive (@liu2014identification), most studies seem to be in between this spectrum (CITE). Particular challenges with the utilization of high-copy number sequences are associated with their rates of divergence.... and... Herein we have resolved major components of the problems of identifying plant material without diagnostic morphological character states using the Angiosperms353 (A353) Hyb-Seq probes (@johnson2019universal), and custom species sequence databases derived via species distribution modelling.

  Our foundation for increasing the quality of metabarcoding results in plants is reducing the number of possible plant species candidates by generating user selected sequence databases. While there are numerous possible approaches for this process, we achieve the selection of possible plant candidate species using digital collections gleaned from Herbaria, (typically Governmental) survey work, and citizen science from a domain exceeding the study area. To these candidate species an approach, such as logistic regression, may be used to identify distances under which taxa are worth further exploring. To these candidate species, we generate species distribution models, which indicate the probability of suitable habitat in a domain, and base the inclusion of these taxa, or representative congeners, upon these results. This approach has the additional benefit of greatly reducing the size of a sequence database, which allows for the usage of genomic size data on personal computers. Moreover, as most next-generation sequence data is deposited as raw-sequence reads, from a processing perspective, it is essential to reduce the candidate species via an approach as such.
  Concomitant to next-generation sequencing has been high-throughput sequencing (cite). Currently the largest plant systematic endeavor ever undertaken, the Kew Plant and Fungal Tree of Life (PAFTOL), is approaching completion (@baker2021exploring). The dataset it creates will contain Hyb-Seq data from at least one species representing each genus in the Plant Kingdom using the popular A353 probes (). These widely available data serve to provide a taxonomically comprehensive backbone for plant metabarcoding. 
  *A brief consideration and example on the inclusion of the temporal dimension of plant flowering, which we suspect will be of importance in sub-tropical, and tropical ecosystems - and others with high seasonality in flowering, is also discussed.* 
  We show that by combining both candidate plant species, and representative sequence data, with only a small amount of new reference sequencing, one is capable of generating ecologically realistic, and biologically relevant results. To test these assertions we utilize a long-term observational study of Bumble Bees, the most important genus of pollinators in the temperate, and one with consistently concerning conservation assessments (@cameron2020global, @goulson2008decline). By generating sequence data from corbiculae, pollen baskets, we seek to determine whether this approach is viable, and if the foraging record of pollen matches the record derived from observations.  **bee stuff here**

# METHODS

## Study System
Observations and sample collection was conducted at The Rocky Mountain Biological Laboratory (RMBL; 38°57.5" N, 106°59.3" W (WGS 84), 2900 m.a.s.l.), Gunnison County, Colorado, USA (see...*APPENDIX 1* for site information). Pollinator observations of *Bombus* Latreille spp. (Apidae Latreille) were conducted from June - August of 2015. The six study sites are in areas characterised by high-montane/subalpine Parkland vegetation communities.

### Spatial Analyses
To develop an ecologically relevant list of vascular plant species, with expected biotic pollination, which may be present at the study sites all records adjacent to the field site were downloaded from BIEN (@bien2022), and these taxa had Species Distribution Models generated to infer their suitability. This list of species served as a reference for which species to include in the genomic sequence databases.

In order to minimise the number of species for which SDM’s were to be generated, BIEN was queried at a distance of up to 100km from our field site and all plant species records were downloaded. In order to estimate the stochasticity of collections, this dataset was bootstrap re-sampled 250 times, with 90% of samples selected, to create a testing dataset. The median of the logistic regression assessing the probability of occurrence of a species record as a function of distance from the study area was used as a threshold distance to include species for distribution modelling.

Species had all records from BIEN within a 50km border of the Omernik level 3 ecoregion which the site is located in (No. 21' Southern Rockies'), downloaded (n = 23,919), (@omernik1987ecoregions). These records were split into two, initially identical, sets, one for generating machine learning models, and the other for Generalised Linear (GLM) and Generalized Additive Models (GAM). The set for generating GLM and GAM records was thinned to reduce spatial autocorrelation in the dataset, as measured by Morans Index (@moran1950notes, @spdep). To both datasets an additional 4029 plots collected from a random stratification of 19% of the land cover in the area of analysis were searched to create true absences (BLM CITATION- need appropriate format for journal). To achieve a larger absence dataset 1000 pseudo-absence records were generated for each taxon, each of which was greater than 10km from an occurrence record. For ML models, these pseudo-absences were reduced so that the number of presence to absence records were 1:1; to achieve this, absence records inside of 10% of the mean sample value, of the presence records, for any predictor were removed.

Species abiotic niche predictors were 26 variables at 30m resolution, for linear regression models these predictors underwent both vifstep (theta = 10, max observations = 12,500) and vifcor (theta = 0.7, max observations = 12,500), and collinear features were removed leaving 16 variables (@usdm2014) (*APPENDIX 6*).

Modelling: Random Forest and Boosted Regression Trees, were sub sampled with 30% test and two replicates each before weighted ensemble based on True Skill Statistics (tss) (@sdmPackage). Generalised linear models (GLM) and Generalised additive models (GAM) with 30% sub sampling and three replicates each were also ensembled using the tss (@sdmPackage). The results of these models were extracted to a polygon feature derived from a minimum-spanning tree which encompasses the study area, and species from either ensemble with greater than 50% habitat suitability were considered present for further purposes (@prim1957minimum). To evaluate these results GBIF (n = 61789) records which had not been assimilated to the BIEN database yet were evaluated using logistic regression (GBIF citation).

535 species were modelled using Generalized Linear Models and Generalized Additive Models. 534 species were modelled using Random Forest and Boosted Regression Trees. *Of the species modelled % were accurately classified as being present at the field station, while % were accurately classified as being absent from the field station* TABLE XX* for more results.*

To evaluate the accuracy of the Species distribution models, additional presence records from GBIF (n = 61789), and AIM (n = 12730) were used as test and training sets (n = 74,519) (CITE AIM AND GBIF). Additional novel absence records were generated from the AIM dataset to create a dataset where each species has balanced presence and absences. 11 or more paired presence and absence records were required for this testing, resulting in 334 species being included in the logistic regression (median = 110.0, $\bar{x}$ = 223.1, max = 1568 record pairs used) with a 70% test split (@caret).

## Morphological Pollen identification
To develop a reference library of  pollen grains which may be present in corbiculae loads, an image reference collection of fuchsin-jelly stained (@beattie1971technique) slides was assembled from slides previously prepared by the authors (n = 21), and other researchers (n = 38) (@brosi2013single). Using 5 years of observational data on *Bombus* Queen Bee foraging at these studies sites (Ogilvie unpublished), as well as the Vascular Plant Checklist (FRASER BUCK 2007), an additional 62 voucher slides were prepared and imaged at 400x (Leica DMLB, Leica MC170 HD Camera, Leica Application Suite V. 4.13.0) from non accessioned herbarium collections to supplement the number of species and clades covered (Appendix 3).  
  In order to determine which plant taxa were distinguishable via light microscopy, and to develop a dichotomous key to pollen morphotypes, Divisive Hierarchical Clustering techniques were used. Ten readily discernible categorical traits were collected from each specimen in the image collection. These traits were transformed using Gower distances, and clustered using Divisive Hierarchical clustering techniques (@cluster2022). Using the cluster dendrogram, elbow plot, and heatmaps (@fpc2022), of these results morphological groups of pollen which could not be resolved via microscopy were delineated, and a dichotomous key was prepared (APPENDIX NO.). This key was then used to identify the pollen grains sampled from corbiculae loads to morphotypes in a consistent manner. 
  To prepare the pollen slides from corbiculae, all corbiculae loads were broken apart and rolled using dissection needlepoints to increase heterogeneity of samples. Cerca 0.5mm2 of pollen was placed onto a ~4mm2 fuchsin jelly cube (@beattie1971technique) atop a graticulated microscope slide, with 20 transects and 20 rows (400 quadrats) (EMS, Hartfield, PA). The jelly was  melted until pollen grains were homogeneously spread across the microscope slide. Slides were sealed with Canada Balsam (Rublev Colours, Willits, CA); all samples are noted in *APPENDIX 3*.
  To identify the pollen present in corbiculae loads, light microscopy at 400x (Zeiss Axioscope A1) was used. In initial sampling in three transects, each pollen grain was identified to morphotype and counted; an additional two transects were scanned for morphotypes unique to that slide, if either transect contained an unique morphotype than all grains in that transect were also identified and counted. Subsequent to the first round of sampling, non-parametric Species Richness Rarefaction curves (@vegans2022), and non-parametric Species Diversity rarefaction curves were used to assess the completeness of sampling (@inextArticle, @inextPackage). Slides not approaching the asymptote of the rarefaction curve were then re-sampled, and analysed iteratively for up to a total of seven transects *APPENDIX 2*.

## Molecular Lab Work
### Reference Plant Library Generation
Using 5 years of observational data on *Bombus* Queen Bee foraging at these studies sites (Ogilvie unpublished), we identified the plant taxa most frequently visited by Queens across all years. We sequenced the 12 most commonly visited taxa twice using samples from one site within the Gunnison River Drainage and one individual from another population. In addition, for any of these 12 focal species which did not have a congener pair in this filtered sample, we included a congener - or a species from a closely related genus to serve as an outgroup. We also sequenced another 15 abundant taxa commonly visited by *Bombus* workers, based on the aforementioned data set (*APPENDIX 4*).

### Plant Genomic DNA Extraction
Plant genomic DNA was isolated from ~ 1 cm2 of leaf tissue from silica-gel dried or herbarium material using a modified cetyltrimethylammonium (CTAB) protocol (@doylesCTAB) that included  two chloroform washes. DNA was quantified using a Nanodrop 2000 (Thermo Fisher Scientific, Waltham, Massachusetts, USA) and Qubit fluorometer (Thermo Fisher Scientific).

### Pollen Genomic DNA Extraction
Pollen genomic DNA was extracted from corbiculae using a CTAB based protocol modified from Lahlamgiahi et al. and Guertler et al. (2014, 2014). A SDS extraction buffer (350µL , 100mM Tris-HCl, 50 mM EDTA, 50 mM NaCl, 10% SDS v/v., pH 7.5) was added followed by vortexing to allow dissolution of corbiculae. Pollen grains were then macerated with Kontes Pellet Pestles, and the tip of these washed with 130 µL of the SDS extraction buffer, samples were then incubated for 1 hour at 30°C. This was followed by the addition of 10% CTAB solution (450ul, of 20 mM Tris-Cl pH. 8.0, 1.4 M NaCl, 10 mM EDTA pH 7.5, 10% CTAB, 5% PVP, ~85% Deionized water) and RNAse (10 uL of 10 mg/mL) and samples were incubated for 40 minutes at 37°C, on heat block (Multi-Blok, Thermo Fisher Scientific, Waltham Massachusetts) set to 40°C. After 20 minutes incubation, Proteinase K (15 µL of 20mg/ml) and  DTT (12.5 µL of 1M in water) were added, and the samples were further incubated at 60°c for 1 hour. Samples were then incubated overnight at 40°C. 500 µL of Phenol-Chloroform-Isoamyl alcohol (25:24:1) were added, vortexed, and centrifuged at 10,000 rpm for 10 minutes and the aqueous phase was pipetted to a 1.5 ml centrifuge tube.

@@

### Fragmentation, Library Preparation & Target Enrichment
Library preparation was performed using the NEBNext Ultra II FS-DNA Library Prep Kit for Illumina (New England BioLabs, Ipswich, Massachusetts, USA) using slightly modified manufacturers recommendation.  Fragmentation was performed at ½ volume of reagents and ¼ enzyme mix for 40 minutes at 37*C, with an input of 500 ng cleaned DNA. Adapter Ligation and PCR enrichment were performed with ½ volumes, while cleanup of products was performed with ½ volume of SPRI beads (Beckman Coulter, Indianapolis, Indiana, USA) and recommended volumes of 80% v./v. ethanol washes. The exception was the herbarium specimens which were not fragmented and only end repaired, with similar library preparation of all samples. Products were analysed on 4% agarose gels, and a Qubit fluorometer.

Libraries were pooled and enriched with the Angiosperms 353 probe kit V.4 (Arbor Biosciences myBaits Target Sequence Capture Kit) by following the manufacturer's protocol and Brewer et al. 2019. Sequencing was performed using an Illumina mi-Seq with 150-bp end reads, (NUSeq Core, Chicago, Illinois).

## Computational Processes and Analyses. 
### Reference Library Data Processing
Sequences were processed using Trimmomatic, which removed sequence adapters, clipped the first 3 bp, discarding reads less than 36 bp, and removing reads if their average PHRED score dropped beneath 20 over a window of 5 bp (@bolger2014trimmomatic). Contigs were generated using HybPiper using target files created by M353 (@johnson2016hybpiper, @mclay2021new).

###  Sequence Identification
A custom Kraken2 database was created by downloading representative species of each genus indicated as being present in the study area by the spatial analyses from the Sequence Read Archive (SRA) NCBI (@wood2019improved). These sequences were processed in the same manner as our novel sequences before being placed into the database. The Kraken2 database was built using default parameters. Kraken2 was run on sequences using default parameters (*APPENDIX 5*). 
  Following Kraken2, Bracken was used to classify sequences to terminal taxa (@lu2017bracken). Results from both Kraken2 and Bracken, results were reclassified manually to identify terminal taxa. For example, when only a single species of a genus was known in the study area, but our database used a representative of another taxon in the genus, this species was coded as the result. The re-coding of sequences from another representative species for the genus to the sole RMBL representative allowed the identification of XX & % more species.

### Identification of Sequence Matching Loci
A local NCBI database was built using the same processed novel and downloaded sequences (@camacho2009blast).

### Temporal Analyses

To estimate the duration of dates in which plant species were flowering Weibull estimates of all spatially modelled taxa were developed (@belitz2020accuracy). Only BIEN records which occurred in the Omernik Level 4 Ecoregions within 15km of the study area (n = 5, or conditionally 6 if enough records not be found in the nearest 5), and which were from herbarium records were included. To remove temporally irrelevant herbarium records, i.e. material collected during times which flowering is impossible at the study area due to snow cover, the SnowUS dataset (@iler2021conceptual, @tran2019cloud) from 2000-2017 was analyzed for the first three days of contiguous snow absence, and the first three days of contiguous snow cover in Fall. Herbarium records after the 3rd quantile for melt, and the 1st quantile for snow cover of these metrics were removed. Species with > 10 records had their weibull distributions estimated for the date when 10% of individuals had begun flowering, when 50% were flowering, and when 90% of individuals had flowered.

# Results

## Microscopic Pollen identification
**Using the fuchsin jelly preparation and light microscopic analyses of grains and scoring of 12 character states resulted in the establishment of XX morphotypes which grains could be reliably classified into. *APPENDIX 7* **. XX Samples were counted and based on rarefaction had over % of expected morphotypes found. The relative abundance of pollen grains in each sample (max % of any species, mean % of all species, min % trace amounts detected).

## Metabarcoding Pollen identification
Kraken2 was able to identify the species richness of pollen samples ($\bar{x}$ = , min = , max = ). Bracken was able to estimate the relative abundance of pollen grains in each sample (max % of any species, $\bar{x}$ % of all species, min % trace amounts detected).

## Spatial Analyses

>>>>

---
title: Plant Metagenomic Barcoding of Pollen Loads Offers Insights on the Foraging
  Patterns of Queen Bumble Bees in the Southern Rocky Mountains, U.S.A.
author: |
  | Reed Clark Benkendorf^[Correspondence: reedbenkendorf2021@u.northwestern.edu] $^1$$^,$$^2$, Jane E. Ogilvie $^3$, Emily J. Woodworth $^1$$^,$$^2$, Sophie Taddeo $^1$$^,$$^2$, Paul J. CaraDonna $^1$$^,$$^2$$^,$$^3$, Jeremie B. Fant $^1$$^,$$^2$ | $^1$Chicago Botanic Garden, 1000 Lake Cook Road, Glencoe, Illinois 60022 USA,  $^2$Plant Biology and Conservation, Northwestern University, Evanston, IL 60208, USA, $^3$ Rocky Mountain Biological Laboratory, P.O. Box 519, Crested Butte, CO 81224, USA
abstract: |
  an abstract will be written to fill this space.
output:
  pdf_document: default
  toc: no
csl: ../citations/methods-in-ecology-and-evolution.csl
bibliography: ../citations/citations.bib
link-citations: yes
fig_caption: yes
header-includes:
- \usepackage{endfloat}
- \usepackage{setspace}\doublespacing
- \usepackage{lineno}
- \linenumbers # cannot co-exist with two columns. 
#classoption: 
#- twocolumn #this makes everything take up two columns, cannot exist with line numbers
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```

@@

# 1 | INTRODUCTION

The inability to reliably identify plants to the level of species often leaves our understanding of ecosystem function and interactions wanting. Current methods to ameliorate this situation include: ignoring these ecologically relevant levels of detail, revisiting plots as diagnostic material becomes temporally available, assistance from taxonomic specialists, or the use of barcoding or other molecular techniques. These approaches are untenable in light of the benefits offered by: species in several morphologically difficult genera which serve as bioindicators, preferred partners in ecological interactions, as well as an increasing lack of taxonomic experts (CITE). Many genera, especially with the formalized advent of integrative taxonomy, have species which are well defined based upon ecological and behavioral rather than morphological properties, the identification of these taxa in degraded areas or without their mutualistic partners is fraught with difficulty. Hindering an understanding of the breadth of habitat which some species occupy, and the interactions they have with other species. 
  The identification of many plant species to terminal taxon is an essential component of nearly all land management programs, where many species in the same genus  (e.g. Sagebrush - *Artemisia* L., Willows - *Salix* L., and Sedges - *Carex* L.) serve as bioindicators (respectively for 'rangelands', streams, and wetlands), as well as in academic research (@Gage2013HistoricalRO, AIM). This endeavour is often mired by lack of diagnostic characters (e.g. flowers, fruits, or roots),  and increasingly the description of cryptic species (@janzen2017nuclear, @oliver2009cryptic). Solutions to this problem are wanting, certain programmes have relied increasingly upon revisiting field sites to identify material using morphological or chemical approaches, whereas academic research has often used high copy number plastid genes as barcodes (Rosentreter et al. 2021, MORE MORE). However, both approaches have significant downsides, the former resource intensive at the landscape scale - and often does not work, while the latter seldom works due to a lack of variability in the currently available barcodes (@liu2014identification).
  Recently barcoding, and metabarcoding, have shown considerable promise in all Kingdoms of life. For example …. . With plants the identification of members of certain clades has been quite successful, whereas with others results have been elusive (@liu2014identification), most applications lay along this spectrum (CITE). Particular challenges with the utilization of high-copy number sequences are associated with their rates of divergence.... and... Herein we have resolved major components of the problems of identifying plant material without diagnostic morphological character states using the Angiosperms353 (A353) Hyb-Seq probes (@johnson2019universal), and custom species sequence databases derived via species distribution modelling, and temporal filtering.

  Our foundation for increasing the quality of metabarcoding results in plants is reducing the number of possible plant species candidates by generating user selected sequence databases relevant to the spatial extent of the study region. While there are numerous possible approaches for this process, we achieve the selection of possible plant candidate species using digital collections gleaned from herbaria, (typically Government) survey work, and citizen science (e.g. iNaturalist), from a domain exceeding the study area. To these candidate species modelling approaches, such as logistic regression, may be used to identify distances under which taxa warrant further exploration. To these candidate species, we generate species distribution models (SDM's), which indicate the probability of suitable habitat in a domain, and base the inclusion of these taxa, or representative congeners, upon these results. This approach has the additional benefit of greatly reducing the size of a sequence database, which allows for the usage of genomic size data on personal computers. Moreover, as most next-generation sequence data is deposited as raw-sequence reads, from a processing perspective, it is essential to reduce the candidate species via an approach as such.
  Currently the largest plant systematic endeavor ever undertaken, the Kew Plant and Fungal Tree of Life (PAFTOL), is approaching completion (@baker2021PAFTOL). This dataset will contain Hyb-Seq data from at least one species representing each genus in the plant kingdom using the popular A353 probes (@baker2021exploring), resulting in over 14,000 represented species. These publically available data serve to provide a taxonomically comprehensive backbone for plant metabarcoding. Data from the 10kP project, which seeks to develop reference genomes from a phylogenetically diverse suite of plants will contribute many more records upon it's intended completion, now slated to be by 2030, similar projects which seek to sequence high amounts of genomes in regions e.g. the 'Darwin Tree of Life' are being undertaken which will contribute data for applicable to enormous spatial domains (@cheng2018tenkp, @darwin2022project, @lewin2022biogenome).

  *A brief consideration and example on the inclusion of the temporal dimension of plant flowering, which we suspect will be of importance in sub-tropical, and tropical ecosystems - and others with high seasonality in flowering, is also discussed.* 
  We show that by combining both candidate plant species, and representative sequence data, with only a small amount of new reference sequencing, one is capable of generating ecologically realistic, and biologically relevant results. To test these assertions we utilize a long-term observational study of Bumble Bees, the most important genus of pollinators in the temperate, and one with consistently concerning conservation assessments (@cameron2020global, @goulson2008decline). By generating sequence data from corbiculae, pollen baskets, we seek to determine whether this approach is viable, and if the foraging record of pollen matches the record derived from observations.  **bee stuff here**

# 2 | METHODS

## Study System & Field Work

Observations and sample collection was conducted at The Rocky Mountain Biological Laboratory (RMBL; 38°57.5" N, 106°59.3" W (WGS 84), 2900 m.a.s.l.), Gunnison County, Colorado, USA (*APPENDIX 1* for site information). Pollinator observations of *Bombus* Latreille spp. (Apidae Latreille) were conducted from June - August of 2015. Observations of *Bombus* foraging took place for one hour at each field site in three 100m transects, where all flowers were also counted and place into abundance bins. Corbiculae loads were,  non-lethally, collected once from all Queen individuals encountered. The six study sites are in areas characterised by high-montane/subalpine Parkland vegetation communities.

### 2.1 | Spatial Analyses

#### 2.1.1 Candidate Species

To develop an ecologically relevant list of vascular plant species, with expected biotic pollination, which may be present at the study sites all records adjacent to the field site were downloaded from the Botanical Information and Ecology Network 'BIEN' (@bien2022), and these taxa had species distribution Dodels (SDMs) generated to predict their suitability. The predicted plant species served as a reference for which species to include in the genomic sequence databases.

In order to minimise the number of species for which SDM’s were to be generated, BIEN was queried at a distance of up to 100km from our field site and all plant species records were downloaded. In order to emulate the perceived stochasticity of collections, this dataset was bootstrap re-sampled 250 times, with 90% of samples selected, to create a testing dataset. The median of the logistic regression assessing the probability of occurrence of a species record as a function of distance from the study area was used as a threshold distance, under which, to include species as candidates for distribution modelling.

#### 2.1.2 Distribution Modelling

Species had all records from BIEN within a 50km border of the Omernik level 3 ecoregion which the site is located in *(No. 21 "Southern Rockies")*, downloaded (n = 23,919) (@omernik1987ecoregions). These records were copied into two, initially identical, sets, one for generating machine learning models (Random Forest, and Boosted Regression Tree's), and the other for Generalised Linear (GLM) and Generalized Additive Models (GAM). The set for generating GLM and GAM records was thinned to reduce spatial autocorrelation in the dataset, as measured by Morans Index (@moran1950notes, @spdep). To both datasets an additional 4029 plots collected from a random stratification of 19% of the land cover in the area of analysis were searched to create true absences (BLM CITATION - need appropriate format for journal). To achieve a larger absence dataset 1000 pseudo-absence records were generated for each taxon, each of which was greater than 10km from an occurrence record. For ML models, these pseudo-absences were reduced so that the ratio of presence to absence records were balanced. To achieve this, absence records inside of 10% of the mean sample value of the presence records, for any predictor were removed; the required number of absence records were then randomly sampled.

Species abiotic niche predictors were 26 variables at 30m resolution, six related to climate,  five soil, four topographic, four related to cloud cover, with the remaining reflecting assorted abiotic parameters (@wilson2016remotely, @wang2016locally, @hengl2017soilgrids250m, @robinson2014earthenv) (*APPENDIX 6*). For linear regression models these predictors underwent both vifstep (theta = 10, max observations = 12,500) and vifcor (theta = 0.7, max observations = 12,500), and collinear features were removed leaving 16 variables (@usdm2014).

Modelling: Random Forest and Boosted Regression Trees, were sub sampled with 30% test and two replicates each before weighted ensemble based on True Skill Statistics (tss) (@sdmPackage). Generalised linear models (GLM) and Generalised additive models (GAM) with 30% sub sampling and three replicates each were also ensembled using the tss (@sdmPackage). The results of these models were extracted to a polygon feature derived from a minimum-spanning tree which encompasses the study area, and species from either ensemble with greater than 50% habitat suitability were considered present for further purposes (@prim1957minimum).

535 species were modelled using Generalized Linear Models and Generalized Additive Models. 534 species were modelled using Random Forest and Boosted Regression Trees.  To evaluate the accuracy of the species distribution models, additional presence records from GBIF (n = 61,789), and AIM (n = 12,730) were used as test and training sets (n = 74,519) for logistic regression (CITE AIM AND GBIF). Additional novel absence records were generated from the AIM dataset to create a dataset where each species has balanced presence and absences. 11 or more paired presence and absence records were required for this testing, resulting in 334 species being included in the logistic regression (median = 110.0, $\bar{x}$ = 223.1, max = 1568 record pairs used) with a 70% test split (@caret).

## 2.2 | Molecular Lab Work

All lab work was carried out at The Daniel F. and Ada L. Rice Plant Conservation Science Center at the Chicago Botanic Garden, Glencoe, Illinois.

#### 2.2.1 | Reference Plant Library Generation

Using 5 years of observational data on *Bombus* Queen Bee foraging at these studies sites, we identified the plant taxa most frequently visited by Queens across all years. We sequenced the 12 most commonly visited taxa twice using samples from one site within the Gunnison River Drainage and one individual from another population. In addition, for any of these 12 focal species which did not have a congener pair in this filtered sample, we included a congener - or a species from a closely related genus to serve as an outgroup. We also sequenced another 15 abundant taxa commonly visited by *Bombus* workers, based on the aforementioned data set (*APPENDIX 4*).

#### 2.2.2 | Plant Genomic DNA Extraction

Plant genomic DNA was isolated from ~ 1 cm2 of leaf tissue from silica-gel dried or herbarium material using a modified cetyltrimethylammonium (CTAB) protocol (@doylesCTAB) that included  two chloroform washes. DNA was quantified using a Nanodrop 2000 (Thermo Fisher Scientific, Waltham, Massachusetts, USA) and Qubit fluorometer (Thermo Fisher Scientific).

#### 2.2.3 | Pollen Genomic DNA Extraction

Pollen genomic DNA was extracted from corbiculae using a CTAB based protocol modified from Lahlamgiahi et al. and Guertler et al. (2014, 2014). A SDS extraction buffer (350µL , 100mM Tris-HCl, 50 mM EDTA, 50 mM NaCl, 10% SDS v/v., pH 7.5) was added followed by vortexing to allow dissolution of corbiculae. Pollen grains were then macerated with Kontes Pellet Pestles, and the tip of these washed with 130 µL of the SDS extraction buffer, samples were then incubated for 1 hour at 30°C. This was followed by the addition of 10% CTAB solution (450ul, of 20 mM Tris-Cl pH. 8.0, 1.4 M NaCl, 10 mM EDTA pH 7.5, 10% CTAB, 5% PVP, ~85% Deionized water) and RNAse (10 uL of 10 mg/mL) and samples were incubated for 40 minutes at 37°C, on heat block (Multi-Blok, Thermo Fisher Scientific, Waltham Massachusetts) set to 40°C. After 20 minutes incubation, Proteinase K (15 µL of 20mg/ml) and  DTT (12.5 µL of 1M in water) were added, and the samples were further incubated at 60°c for 1 hour. Samples were then incubated overnight at 40°C. 500 µL of Phenol-Chloroform-Isoamyl alcohol (25:24:1) were added, vortexed, and centrifuged at 10,000 rpm for 10 minutes and the aqueous phase was pipetted to a 1.5 ml centrifuge tube.

@@

#### 2.2.4 | Fragmentation, Library Preparation & Target Enrichment

Library preparation was performed using the NEBNext Ultra II FS-DNA Library Prep Kit for Illumina (New England BioLabs, Ipswich, Massachusetts, USA) using slightly modified manufacturers recommendation.  Fragmentation was performed at ½ volume of reagents and ¼ enzyme mix for 40 minutes at 37*C, with an input of 500 ng cleaned DNA. Adapter Ligation and PCR enrichment were performed with ½ volumes, while cleanup of products was performed with ½ volume of SPRI beads (Beckman Coulter, Indianapolis, Indiana, USA) and recommended volumes of 80% v./v. ethanol washes. The exception was the herbarium specimens which were not fragmented and only end repaired, with similar library preparation of all samples. Products were analysed on 4% agarose gels, and a Qubit fluorometer. 
Libraries were pooled and enriched with the Angiosperms 353 probe kit V.4 (Arbor Biosciences myBaits Target Sequence Capture Kit) by following the manufacturer's protocol and Brewer et al. 2019. Sequencing was performed using an Illumina mi-Seq with 150-bp end reads, (NUSeq Core, Chicago, Illinois).

###  2.2.5 | Computational Processes and Analyses. 
####  2.2.5.1 | Reference Library Data Processing

Sequences were processed using Trimmomatic, which removed sequence adapters, clipped the first 3 bp, discarding reads less than 36 bp, and removing reads if their average PHRED score dropped beneath 20 over a window of 5 bp (@bolger2014trimmomatic). Contigs were generated using HybPiper using target files created by M353 (@johnson2016hybpiper, @mclay2021new).

#### 2.2.5.2 | Sequence Identification

A custom Kraken2 database was created by downloading representative species of each genus indicated as being present in the study area by the spatial analyses from the Sequence Read Archive (SRA) NCBI (@wood2019improved). These sequences were processed in the same manner as our novel sequences before being placed into the database. The Kraken2 database was built using default parameters. Kraken2 was run on sequences using default parameters (*APPENDIX 5*). 
  Following Kraken2, Bracken was used to classify sequences to terminal taxa (@lu2017bracken). Results from both Kraken2 and Bracken, results were reclassified manually to identify terminal taxa. For example, when only a single species of a genus was known in the study area, but our database used a representative of another taxon in the genus, this species was coded as the result. The re-coding of sequences from another representative species for the genus to the sole RMBL representative allowed the identification of XX & % more species.

#### 2.2.5.3 | Identification of Sequence Matching Loci

A local NCBI database was built using the same processed novel and downloaded sequences (@camacho2009blast).

## 2.2.5.4 | Morphological Pollen identification

To develop a reference library of  pollen grains which may be present in corbiculae loads, an image reference collection of fuchsin-jelly stained (@beattie1971technique) slides was assembled from slides previously prepared by the authors (n = 21), and other researchers (n = 38) (@brosi2013single). Using 5 years of observational data on *Bombus* Queen Bee foraging at these studies sites (Ogilvie unpublished), as well as the Vascular Plant Checklist (@fraser2007vpc), an additional 62 voucher slides for species were prepared and imaged at 400x (Leica DMLB, Leica MC170 HD Camera, Leica Application Suite V. 4.13.0) from non accessioned herbarium collections to supplement the number of species and clades covered (Appendix 3).  
  In order to determine which plant taxa were distinguishable via light microscopy, and to develop a dichotomous key to pollen morphotypes, Divisive Hierarchical Clustering techniques were used. Ten readily discernible categorical traits were collected from each specimen in the image collection. These traits were transformed using Gower distances, and clustered using Divisive Hierarchical clustering techniques (@cluster2022). Using the cluster dendrogram, elbow plot, and heatmaps (@fpc2022), of these results morphological groups of pollen which could not be resolved via microscopy were delineated, and a dichotomous key was prepared (APPENDIX NO.). This key was then used to identify the pollen grains sampled from corbiculae loads to morphotypes in a consistent manner. 
  To prepare the pollen slides from corbiculae, all corbiculae loads were broken apart and rolled using dissection needlepoints to increase heterogeneity of samples. *Cerca* 0.5mm^2 of pollen was placed onto a ~4mm^2 fuchsin jelly cube (@beattie1971technique) atop a graticulated microscope slide, with 20 transects and 20 rows (400 quadrats) (EMS, Hartfield, PA). The jelly was melted, with stirring, until pollen grains were homogeneously spread across the microscope slide. Slides were sealed with Canada Balsam (Rublev Colours, Willits, CA) followed by sealing with nail polish; all samples are noted in *APPENDIX 3*.
  To identify the pollen present in corbiculae loads, light microscopy at 400x (Zeiss Axioscope A1) was used. In initial sampling in three transects, each pollen grain was identified to morphotype and counted; an additional two transects were scanned for morphotypes unique to that slide, if either transect contained an unique morphotype than all grains in that transect were also identified and counted. Subsequent to the first round of sampling, non-parametric species richness rarefaction curves (@vegans2022), and non-parametric species diversity rarefaction curves were used to assess the completeness of sampling (@inextArticle, @inextPackage). Slides not approaching the asymptote of the rarefaction curve were then re-sampled, and analysed iteratively for up to a total of seven transects *APPENDIX 2*.
